---
title: "COVID Analysis: Philippines vs Japan"
author: "K. Chu"
date: "9/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'm from the Philippines, and I got curious how the situation looks like. 
However, I have moved to Japan for work, and I'm also curious what the situation
looks like here. Am I safer here in Japan, or should I go back home to the
Philippines?

```{r import-data}
library(tidyverse)
library(lubridate)
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

filenames <- c("time_series_covid19_confirmed_global.csv",
               "time_series_covid19_deaths_global.csv",
               "time_series_covid19_recovered_global.csv")
urls <- str_c(url_in, filenames)
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
global_recoveries <- read_csv(urls[3])

summary(global_cases)

# the tables appear to be very wide; show summary for only global_cases
# as the structures of global_deaths and global_recoveries are similar

# summary(global_deaths)
# summary(global_recoveries)

# rename the columns with slashes
global_cases <- global_cases %>% 
  rename(Country_Region=`Country/Region`,
         Province_State=`Province/State`) 

global_deaths <- global_deaths %>% 
  rename(Country_Region=`Country/Region`,
         Province_State=`Province/State`) 

global_recoveries <- global_recoveries %>% 
  rename(Country_Region=`Country/Region`,
         Province_State=`Province/State`) 
```

I am getting my data from the Center for Systems Science and Engineering (CSSE)
at Johns Hopkins University. They have a GitHub repository for global cases, 
deaths, and recoveries, so I will be using this repository.

The three files here contain a lot of information, but they show a tally of 
cases or deaths or recoveries per country per day, with each day being a new 
column. We have a very wide table. Since I want to more easily see the number of
cases, deaths and recoveries, I will have to transform the data.

Also, since these datasets create wide tables, excuse me as I show the summary 
of only one of them. Showing the summaries of the other two would take up a lot 
of space. The other two datasets have a similar structure, with deaths and 
recoveries taking the place of cases in their respective datasets.

However, for recoveries, it appears that we don't have recorded data past August
4, 2021, as we get 0s from August 5, 2021 onward. I will also have to address 
this in my transformation.

```{r set-recoveries-last-available-date}
recoveries_last_available_date <- "2021-08-04"
```

```{r get-ph-data}
# get philippine cases
ph_cases <- global_cases %>% 
  filter(Country_Region == "Philippines") %>%
  pivot_longer(cols= -c(Province_State,
                        Country_Region,
                        Lat,
                        Long),
               names_to="date",
               values_to="cases") %>%
  select(-c(Lat, Long))

summary(ph_cases)

# get philippine deaths
ph_deaths <- global_deaths %>% 
  filter(Country_Region == "Philippines") %>%
  pivot_longer(cols= -c(Province_State,
                        Country_Region,
                        Lat,
                        Long),
               names_to="date",
               values_to="deaths") %>%
  select(-c(Lat, Long))

summary(ph_deaths)

# get philippine recoveries
ph_recoveries <- global_recoveries %>% 
  filter(Country_Region == "Philippines") %>%
  pivot_longer(cols= -c(Province_State,
                        Country_Region,
                        Lat,
                        Long),
               names_to="date",
               values_to="recoveries") %>%
  select(-c(Lat, Long))

summary(ph_recoveries)

# combine them into ph_data
ph_data <- ph_cases %>%
  full_join(ph_deaths) %>%
  full_join(ph_recoveries) %>%
  mutate(date = mdy(date)) %>%
  filter(cases > 0)

# mark the recoveries as NA past the last available date for recoveries
ph_data$recoveries <- ifelse(ph_data$date > as.Date(recoveries_last_available_date),
                             NA,
                             ph_data$recoveries)

summary(ph_data)

```

Here, I filtered out the data for the Philippines. I also transformed the tables
using pivot_longer so that I can more easily visualize the data. I joined the
two tables together so that I can pair the number of cases with the number of 
deaths. I'm also more interested in seeing things after the Philippines started 
having cases, so I only retained the rows that had at least one case.

As of this writing, the latest entry was for September 22, 2021, with the number
of total cases over time being 2,417,419 and the number of total deaths being
37,228. The number of recoveries, as of August 4, 2021, is 1,528,422.

```{r get-jp-data}
# get japan cases
jp_cases <- global_cases %>% 
  filter(Country_Region == "Japan") %>%
  pivot_longer(cols= -c(Province_State,
                        Country_Region,
                        Lat,
                        Long),
               names_to="date",
               values_to="cases") %>%
  select(-c(Lat, Long))

summary(jp_cases)

# get japan deaths
jp_deaths <- global_deaths %>% 
  filter(Country_Region == "Japan") %>%
  pivot_longer(cols= -c(Province_State,
                        Country_Region,
                        Lat,
                        Long),
               names_to="date",
               values_to="deaths") %>%
  select(-c(Lat, Long))

summary(jp_deaths)

# get japan recoveries
jp_recoveries <- global_recoveries %>% 
  filter(Country_Region == "Japan") %>%
  pivot_longer(cols= -c(Province_State,
                        Country_Region,
                        Lat,
                        Long),
               names_to="date",
               values_to="recoveries") %>%
  select(-c(Lat, Long))

summary(jp_recoveries)

# combine them all into jp_data
jp_data <- jp_cases %>%
  full_join(jp_deaths) %>%
  full_join(jp_recoveries) %>%
  mutate(date = mdy(date)) %>%
  filter(cases > 0)

# mark the recoveries as NA past the last available date for recoveries
jp_data$recoveries <- ifelse(jp_data$date > as.Date(recoveries_last_available_date),
                             NA,
                             jp_data$recoveries)

summary(jp_data)

```

I did the same transformation for the Japanese data. The numbers look better
here, compared to the Philippines. The number of total cases over time is
1,685,325 as of September 22, 2021, and the number of total deaths over time
is 17,348.

The number of recoveries, as of August 4, 2021, is 852,451. This might appear 
low, but given that the number of cases is relatively low compared to the 
Philippines, I think that this wouldn't be unexpected.

```{r visualize-ph-cases-and-deaths}

ph_data %>%
  ggplot(aes(x=date, y=cases)) + 
  geom_line(aes(color="Cases")) + 
  geom_point(aes(color="Cases")) + 
  
  geom_line(aes(y=deaths, color="Deaths")) + 
  geom_point(aes(y=deaths, color="Deaths")) +
  
  geom_line(aes(y=recoveries, color="Recoveries")) + 
  geom_point(aes(y=recoveries, color="Recoveries")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="COVID19 Cases, Deaths, and Recoveries in the Philippines", 
       x="Date",
       y=NULL)

```

```{r visualize-jp-cases-and-deaths}

jp_data %>%
  ggplot(aes(x=date, y=cases)) + 
  geom_line(aes(color="Cases")) + 
  geom_point(aes(color="Cases")) + 
  
  geom_line(aes(y=deaths, color="Deaths")) + 
  geom_point(aes(y=deaths, color="Deaths")) +
  
  geom_line(aes(y=recoveries, color="Recoveries")) + 
  geom_point(aes(y=recoveries, color="Recoveries")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="COVID19 Cases, Deaths, and Recoveries in Japan", 
       x="Date",
       y=NULL)

```

The graphs for both countries appear similar, although there seemed to be a time
where the cases and deaths for the Philippines stagnated in early 2020. Let's
see how the two countries measure up when I include their cases in the same 
graph and do the same for their deaths and their recoveries.

It appears that in both countries, the number of recoveries closely follows the
number of cases. This is quite good news.

```{r visualize-cases-for-both}

ph_data %>%
  ggplot(aes(x=date, y=cases)) + 
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) + 
  
  geom_line(data=jp_data, aes(color="Japan")) +
  geom_point(data=jp_data, aes(color="Japan")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="COVID19 Cases: Philippines vs Japan", 
       x="Date",
       y=NULL)

```

For this graph, we can see that the Philippines had a lower case count at the
start, but it soon overtook Japan. There might have been something to do with 
the early stagnation of cases. Why was there a stagnation here?

```{r visualize-deaths-for-both}

ph_data %>%
  ggplot(aes(x=date, y=deaths)) + 
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) + 
  
  geom_line(data=jp_data, aes(color="Japan")) +
  geom_point(data=jp_data, aes(color="Japan")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="COVID19 Deaths: Philippines vs Japan", 
       x="Date",
       y=NULL)

```

The same trend could be seen in this graph for deaths. However, the Philippines
overtook Japan in the number of deaths much earlier in 2020.

```{r visualize-recoveries-for-both}

ph_data %>%
  ggplot(aes(x=date, y=recoveries)) + 
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) + 
  
  geom_line(data=jp_data, aes(color="Japan")) +
  geom_point(data=jp_data, aes(color="Japan")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="COVID19 Recoveries: Philippines vs Japan", 
       x="Date",
       y=NULL)

```

The number of recoveries for the Philippines overtook that of Japan around July
2020. This was to be expected since the Philippines had a higher number of cases
to begin with.

```{r add-ph-new-cases-deaths-recoveries-columns}
# add columns for new cases, new deaths, and new recoveries
ph_data <- ph_data %>%
  mutate(new_cases=cases - lag(cases),
         new_deaths=deaths - lag(deaths),
         new_recoveries=recoveries - lag(recoveries))

ph_data$new_cases <- ifelse(is.na(ph_data$new_cases), 
                            ph_data$cases,
                            ph_data$new_cases)

ph_data$new_deaths <- ifelse(is.na(ph_data$new_deaths), 
                             ph_data$deaths,
                             ph_data$new_deaths)

ph_data$new_recoveries <- ifelse(is.na(ph_data$new_recoveries), 
                                 ph_data$recoveries,
                                 ph_data$new_recoveries)

ph_data %>%
  ggplot(aes(x=date, y=new_cases)) + 
  geom_line(aes(color="New Cases")) + 
  geom_point(aes(color="New Cases")) + 
  
  geom_line(aes(y=new_deaths, color="New Deaths")) + 
  geom_point(aes(y=new_deaths, color="New Deaths")) +
  
  geom_line(aes(y=new_recoveries, color="New Recoveries")) + 
  geom_point(aes(y=new_recoveries, color="New Recoveries")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="New Daily Cases, Deaths, and Recoveries of COVID19 in the Philippines", 
       x="Date",
       y=NULL)
```

I thought of adding columns for new cases, new deaths, and new recoveries. I 
then wanted to see how these looked like over time.

There appears to be an upward trend, for all three new columns, but the number 
of new deaths and the number of new recoveries per day appear to 
vary wildly. Still, I can see an upward trend.

The number of new cases per day appears to be more stable compared to the number
of new deaths and the number of new recoveries.

```{r add-jp-new-cases-and-deaths-columns}
# add columns for new cases, new deaths, and new recoveries
jp_data <- jp_data %>%
  mutate(new_cases=cases - lag(cases),
         new_deaths=deaths - lag(deaths),
         new_recoveries=recoveries - lag(recoveries))

jp_data$new_cases <- ifelse(is.na(jp_data$new_cases), 
                            jp_data$cases,
                            jp_data$new_cases)

jp_data$new_deaths <- ifelse(is.na(jp_data$new_deaths), 
                             jp_data$deaths,
                             jp_data$new_deaths)

jp_data$new_recoveries <- ifelse(is.na(jp_data$new_recoveries), 
                                 jp_data$recoveries,
                                 jp_data$new_recoveries)

jp_data %>%
  ggplot(aes(x=date, y=new_cases)) + 
  geom_line(aes(color="New Cases")) + 
  geom_point(aes(color="New Cases")) + 
  
  geom_line(aes(y=new_deaths, color="New Deaths")) + 
  geom_point(aes(y=new_deaths, color="New Deaths")) +
  
  geom_line(aes(y=new_recoveries, color="New Recoveries")) + 
  geom_point(aes(y=new_recoveries, color="New Recoveries")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="New Daily Cases, Deaths, and Recoveries of COVID19 in Japan", 
       x="Date",
       y=NULL)
```

Japan's case appears to be much different. The number of new cases also appears
to be on an upward trend, with several surges. The number of new recoveries
appears to follow the number of new cases closely, although there are times when
that number dips to zero.

I would like to compare how the two countries measure up with the new columns.

```{r visualize-new-cases-for-both}

ph_data %>%
  ggplot(aes(x=date, y=new_cases)) + 
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) + 
  
  geom_line(data=jp_data, aes(color="Japan")) +
  geom_point(data=jp_data, aes(color="Japan")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="New Daily COVID19 Cases: Philippines vs Japan", 
       x="Date",
       y=NULL)

```

In general, Japan tends to have fewer new daily cases than the Philippines. The
Philippines might have fewer cases from time to time, but these periods are 
shorter compared to when Japan had more new daily cases.

```{r visualize-new-deaths-for-both}

ph_data %>%
  ggplot(aes(x=date, y=new_deaths)) + 
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) + 
  
  geom_line(data=jp_data, aes(color="Japan")) +
  geom_point(data=jp_data, aes(color="Japan")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="New Daily COVID19 Deaths: Philippines vs Japan", 
       x="Date",
       y=NULL)

```

This graph doesn't look very easy on the eyes. Still, Japan tends to have fewer
new deaths compared to the Philippines.

```{r visualize-new-recoveries-for-both}

ph_data %>%
  ggplot(aes(x=date, y=new_recoveries)) + 
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) + 
  
  geom_line(data=jp_data, aes(color="Japan")) +
  geom_point(data=jp_data, aes(color="Japan")) +
  
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="New Daily COVID19 Recoveries: Philippines vs Japan", 
       x="Date",
       y=NULL)

```

The number of new recoveries for both countries appear to be tied. There is a
period where the number of new recoveries vary wildly, with Japan tending to 
have days where there were no new recoveries while the Philippines tended to 
have days where the number of new recoveries shot up before coming back down.

```{r add-deaths-to-cases-column}
ph_data <- ph_data %>%
  mutate(deaths_to_cases=deaths / cases)

jp_data <- jp_data %>%
  mutate(deaths_to_cases=deaths / cases)
  
```

I think another measure I should look into is the case fatality rate, or the 
number of deaths divided by the number of cases.

```{r visualize-case-fatality-rates}

ph_data %>%
  ggplot(aes(x=date, y=deaths_to_cases)) +
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) +
  
  geom_line(data=jp_data, aes(color="Japan")) +
  geom_point(data=jp_data, aes(color="Japan")) +
  
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="Case Fatality Rate of COVID19: Philippines vs Japan",
       x="Date",
       y=NULL)

```

It didn't look so good for the Philippines at first, but the numbers went back
down. The case fatality rate for both countries appears to have settled to 
almost close to 0. Let's take a closer look at things in 2021.

```{r visualize-case-fatality-rates-2021}

# get data from 2021
ph_data_2021 <- ph_data %>%
  filter(date >= as.Date("2021-01-01"))

jp_data_2021 <- jp_data %>%
  filter(date >= as.Date("2021-01-01"))

ph_data_2021 %>%
  ggplot(aes(x=date, y=deaths_to_cases)) +
  geom_line(aes(color="Philippines")) + 
  geom_point(aes(color="Philippines")) +
  
  geom_line(data=jp_data_2021, aes(color="Japan")) +
  geom_point(data=jp_data_2021, aes(color="Japan")) +
  
  theme(legend.position="bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title="Case Fatality Rate of COVID19 in 2021: Philippines vs Japan",
       x="Date",
       y=NULL)

```

According to this graph, the case fatality rates went way down, below 0.02.

The lower bound for the Philippines appears to be 0.015 while it is 0.010 for
Japan.

This is great since this means that COVID is not particularly fatal. It appears
that for every 100 cases, I can expect around 1 or 2 deaths in the Philippines
while I can expect close to 1 death in Japan.

I would like to create a model to see how things would play out in the future.
I think it's more accurate to use only this year's data with regards to case 
fatality rate since the outbreak was such a sudden shock to the world last year.
I think now that the world has had a chance to stabilize somewhat, this year's
numbers should be more relevant.

```{r model-ph-case-fatality-rate}
mod_ph <- lm(deaths_to_cases ~ date, data=ph_data_2021)
summary(mod_ph)

ph_w_pred <- ph_data_2021 %>% mutate(pred = predict(mod_ph))

ph_w_pred %>% ggplot() +
  geom_point(aes(x=date, y=deaths_to_cases), color="blue") + 
  geom_point(aes(x=date, y=pred), color="red") +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title="Philippine 2021 Case Fatality Rate: Actual vs Model",
       y="Case Fatality Rate",
       x="Date")
  
```

```{r model-jp-case-fatality-rate}
mod_jp <- lm(deaths_to_cases ~ date, data=jp_data_2021)
summary(mod_jp)

jp_w_pred <- jp_data_2021 %>% mutate(pred = predict(mod_jp))

jp_w_pred %>% ggplot() +
  geom_point(aes(x=date, y=deaths_to_cases), color="blue") + 
  geom_point(aes(x=date, y=pred), color="red") +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title="Japan 2021 Case Fatality Rate: Actual vs Model",
       y="Case Fatality Rate",
       x="Date")
```

Both models predict a downward trend for both countries. This is great news, 
especially after taking into consideration that my model predicts that the
ratio of deaths to cases is higher than was observed in September 2021.

# Conclusion

It appears that there is a downward trend in the case fatality rates for both
countries. The rates for both countries appear to continue going down, as 
predicted by my models.

It appears that Japan has a slightly lower case fatality rate as of September
2021. Perhaps it was a good thing for me to have come here, after all. I do miss
seeing my family, however.

I could only look forward to the day that the numbers say zero. However, this
may be impossible, and I have to temper my expectations so.

I have heard that age is a comorbidity when it came to COVID. Given that Japan's
population is aging, I would have expected that Japan's case fatality rate would
have been higher. However, I should also take into consideration that Japan's
healthcare system could be superior compared to the Philippines's.

Focusing solely on raw cases, deaths, and recoveries might not paint an accurate
picture, so I opted to go for a metric like the case fatality rate. This would
control for the different numbers of cases and deaths.

```{r show-session-info}
sessionInfo()
```

