---
title: "NYPD Shooting Incidents by Borough"
author: "K. Chu"
date: "9/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I thought about what would be an interesting question to answer, and I thought
that finding the most dangerous borough and the safest borough would be 
something I wanted to know.

```{r import-data}
library(tidyverse)
library(lubridate)

url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

incidents <- read_csv(url_in)
summary(incidents)
```

I have gotten my data from the City of New York's data website. This is their
shooting incident data, as recorded from 2006 all the way until the end of 2020.

It contains quite a lot of information, such as some information about the
perpetrator and the victim. There are fields for the race, age group, and sex of
both victims and perpetrators. It also includes whether the victim was murdered.

It also includes location information about each shooting, even up to latitude 
and longitude! It even includes a description of the location, as well as the
precinct and jurisdiction code of each location where the shootings take place.

Of course, it also includes the time and date of each shooting.

```{r clean-data}
incidents <-incidents %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),  # convert OCCUR_DATE to date
         BORO = factor(BORO),
         LOCATION_DESC = factor(LOCATION_DESC),
         STATISTICAL_MURDER_FLAG = factor(STATISTICAL_MURDER_FLAG),
         PERP_AGE_GROUP = factor(PERP_AGE_GROUP),
         PERP_SEX = factor(PERP_SEX),
         PERP_RACE = factor(PERP_RACE),
         VIC_AGE_GROUP = factor(VIC_AGE_GROUP),
         VIC_SEX = factor(VIC_SEX),
         VIC_RACE = factor(VIC_RACE)) %>%
  # remove unneeded columns
  select(-c(INCIDENT_KEY, PRECINCT, JURISDICTION_CODE,
            X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat)) %>%
  # replace the NA entries as "unknown" or some variation
  replace_na(list(LOCATION_DESC="NONE",
                  PERP_AGE_GROUP="UNKNOWN",
                  PERP_SEX="U",
                  PERP_RACE="UNKNOWN"))

summary(incidents)
```

There are fields where the entries are blank. Examples include:

- LOCATION_DESC
- PERP_AGE_GROUP
- PERP_SEX
- PERP_RACE

What I did was to change the blank entries to reflect the same value used to 
denote an unknown value.

```{r transform-data}
# sort by date
incidents <- incidents %>%
  arrange(OCCUR_DATE)

# use month instead of specific day
incidents <- incidents %>%
  mutate(Occur_Month = floor_date(OCCUR_DATE, "month"))

incidents$rec <- 1

# create a column for cumulative shootings by borough
with_cumulative_cases <- incidents %>%
  group_by(BORO) %>%
  mutate(Cumulative_Shootings = cumsum(rec)) %>%
  select(-rec) %>%
  ungroup()

# get the number of cumulative shootings per month by borough
with_cumulative_cases <- with_cumulative_cases %>%
  group_by(Occur_Month, BORO) %>%
  top_n(1, Cumulative_Shootings) %>%
  select(c(Occur_Month, BORO, Cumulative_Shootings)) %>%
  ungroup()

# divide the cases by borough
queens <- with_cumulative_cases %>%
  filter(BORO == "QUEENS")

manhattan <- with_cumulative_cases %>%
  filter(BORO == "MANHATTAN")

brooklyn <- with_cumulative_cases %>%
  filter(BORO == "BROOKLYN")

bronx <- with_cumulative_cases %>%
  filter(BORO == "BRONX")

staten_island <- with_cumulative_cases %>%
  filter(BORO == "STATEN ISLAND")
```

Here, I tried grouping the shootings by borough. I also realized that I didn't
quite need information about the victim and the perpetrator, so I simply removed
their columns.

```{r visualize-data}

queens %>%
  ggplot(aes(x = Occur_Month, y = Cumulative_Shootings)) +
  geom_line(aes(color="Queens")) +
  geom_point(aes(color="Queens")) +
  
  geom_line(data=manhattan, aes(color="Manhattan")) +
  geom_point(data=manhattan, aes(color="Manhattan")) +
  
  geom_line(data=brooklyn, aes(color="Brooklyn")) +
  geom_point(data=brooklyn, aes(color="Brooklyn")) +
  
  geom_line(data=bronx, aes(color="Bronx")) +
  geom_point(data=bronx, aes(color="Bronx")) +
  
  geom_line(data=staten_island, aes(color="Staten Island")) +
  geom_point(data=staten_island, aes(color="Staten Island")) +
  
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title="Total Shootings in New York by Borough",
       y="Total shootings",
       x="Month")

```

The graphs show that Bronx and Brooklyn are the places with the most number of
shooting incidents while Staten Island has the least number of shooting
incidents. It has led me to ask if Bronx and Brooklyn truly are the most
dangerous places in New York and if Staten Island is the safest place in 
New York.

```{r add-second-analysis}
# add new shootings column

queens <- queens %>%
  mutate(New_Shootings = Cumulative_Shootings - lag(Cumulative_Shootings))
queens$New_Shootings <- ifelse(is.na(queens$New_Shootings), 
                               queens$Cumulative_Shootings,
                               queens$New_Shootings)

manhattan <- manhattan %>%
  mutate(New_Shootings = Cumulative_Shootings - lag(Cumulative_Shootings))
manhattan$New_Shootings <- ifelse(is.na(manhattan$New_Shootings), 
                               manhattan$Cumulative_Shootings,
                               manhattan$New_Shootings)

brooklyn <- brooklyn %>%
  mutate(New_Shootings = Cumulative_Shootings - lag(Cumulative_Shootings))
brooklyn$New_Shootings <- ifelse(is.na(brooklyn$New_Shootings), 
                               brooklyn$Cumulative_Shootings,
                               brooklyn$New_Shootings)

bronx <- bronx %>%
  mutate(New_Shootings = Cumulative_Shootings - lag(Cumulative_Shootings))
bronx$New_Shootings <- ifelse(is.na(bronx$New_Shootings), 
                               bronx$Cumulative_Shootings,
                               bronx$New_Shootings)

staten_island <- staten_island %>%
  mutate(New_Shootings = Cumulative_Shootings - lag(Cumulative_Shootings))
staten_island$New_Shootings <- ifelse(is.na(staten_island$New_Shootings), 
                               staten_island$Cumulative_Shootings,
                               staten_island$New_Shootings)

```

Here, I added a new column for new shootings. I also made sure that there were
no empty fields. I noticed that the first entry was empty because there was no
value for lag(Cumulative_Shootings) at that point. For this, I simply copied
over the Cumulative_Shootings value for the first month.

```{r visualize-second-analysis}

queens %>%
  ggplot(aes(x = Occur_Month, y = New_Shootings)) +
  geom_line(aes(color="Queens")) +
  geom_point(aes(color="Queens")) +
  
  geom_line(data=manhattan, aes(color="Manhattan")) +
  geom_point(data=manhattan, aes(color="Manhattan")) +
  
  geom_line(data=brooklyn, aes(color="Brooklyn")) +
  geom_point(data=brooklyn, aes(color="Brooklyn")) +
  
  geom_line(data=bronx, aes(color="Bronx")) +
  geom_point(data=bronx, aes(color="Bronx")) +
  
  geom_line(data=staten_island, aes(color="Staten Island")) +
  geom_point(data=staten_island, aes(color="Staten Island")) +
  
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="New Monthly Shootings in New York by Borough",
       y="New Monthly Shootings",
       x="Month")


```

According to this graph, there appear to be several sharp spikes in shooting
incidents in Brooklyn and Bronx. The other three boroughs, Manhattan, Queens,
and Staten Island, appear to be more muted by comparison.

However, there is an even higher surge in shooting incidents in 2020. This 
increase is particularly sharp in Brooklyn and Bronx. What caused this spike?
This sharp increase is unprecedented, especially in Brooklyn.

```{r create-monthly-model-queens}

# queens monthly model
mod_queens <- lm(New_Shootings ~ Occur_Month, data=queens)
summary(mod_queens)

queens_w_pred <- queens %>% mutate(pred = predict(mod_queens))

queens_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Month, y = New_Shootings), color = "blue") + 
  geom_point(aes(x = Occur_Month, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Queens New Monthly Shootings: Actual vs Model",
       y="New Monthly Shootings",
       x="Month")

```

```{r create-monthly-model-manhattan}

# manhattan monthly model
mod_manhattan <- lm(New_Shootings ~ Occur_Month, data=manhattan)
summary(mod_manhattan)

manhattan_w_pred <- manhattan %>% mutate(pred = predict(mod_manhattan))

manhattan_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Month, y = New_Shootings), color = "blue") + 
  geom_point(aes(x = Occur_Month, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Manhattan New Monthly Shootings: Actual vs Model",
       y="New Monthly Shootings",
       x="Month")

```

```{r create-monthly-model-brooklyn}

# brooklyn monthly model
mod_brooklyn <- lm(New_Shootings ~ Occur_Month, data=brooklyn)
summary(mod_brooklyn)

brooklyn_w_pred <- brooklyn %>% mutate(pred = predict(mod_brooklyn))

brooklyn_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Month, y = New_Shootings), color = "blue") + 
  geom_point(aes(x = Occur_Month, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Brooklyn New Monthly Shootings: Actual vs Model",
       y="New Monthly Shootings",
       x="Month")

```

```{r create-monthly-model-bronx}

# bronx monthly model
mod_bronx <- lm(New_Shootings ~ Occur_Month, data=bronx)
summary(mod_bronx)

bronx_w_pred <- bronx %>% mutate(pred = predict(mod_bronx))

bronx_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Month, y = New_Shootings), color = "blue") + 
  geom_point(aes(x = Occur_Month, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Bronx New Monthly Shootings: Actual vs Model",
       y="New Monthly Shootings",
       x="Month")

```

```{r create-monthly-model-staten-island}

# staten island monthly model
mod_staten_island <- lm(New_Shootings ~ Occur_Month, data=staten_island)
summary(mod_staten_island)

staten_island_w_pred <- staten_island %>% mutate(pred = predict(mod_staten_island))

staten_island_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Month, y = New_Shootings), color = "blue") + 
  geom_point(aes(x = Occur_Month, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Staten Island New Monthly Shootings: Actual vs Model",
       y="New Monthly Shootings",
       x="Month")

```

I try to make a linear model for each of the boroughs.

A linear model isn't particularly accurate when trying to predict the number of
shootings per month. How would our graphs look like when we use yearly data
instead?

```{r add-third-analysis}

# queens
queens_by_year <- queens %>%
  mutate(Occur_Year = floor_date(Occur_Month, "year"))

queens_by_year <- queens_by_year %>%
  group_by(Occur_Year, BORO) %>%
  top_n(1, Cumulative_Shootings) %>%
  select(c(Occur_Year, BORO, Cumulative_Shootings)) %>%
  ungroup()

queens_by_year <- queens_by_year %>%
  mutate(Cumulative_Shootings_By_Year = Cumulative_Shootings)

queens_by_year <- queens_by_year %>%
  mutate(New_Shootings_By_Year = Cumulative_Shootings_By_Year - lag(Cumulative_Shootings_By_Year))
queens_by_year$New_Shootings_By_Year <- ifelse(is.na(queens_by_year$New_Shootings_By_Year), 
                                               queens_by_year$Cumulative_Shootings_By_Year,
                                               queens_by_year$New_Shootings_By_Year)

# manhattan
manhattan_by_year <- manhattan %>%
  mutate(Occur_Year = floor_date(Occur_Month, "year"))

manhattan_by_year <- manhattan_by_year %>%
  group_by(Occur_Year, BORO) %>%
  top_n(1, Cumulative_Shootings) %>%
  select(c(Occur_Year, BORO, Cumulative_Shootings)) %>%
  ungroup()

manhattan_by_year <- manhattan_by_year %>%
  mutate(Cumulative_Shootings_By_Year = Cumulative_Shootings)

manhattan_by_year <- manhattan_by_year %>%
  mutate(New_Shootings_By_Year = Cumulative_Shootings_By_Year - lag(Cumulative_Shootings_By_Year))
manhattan_by_year$New_Shootings_By_Year <- ifelse(is.na(manhattan_by_year$New_Shootings_By_Year), 
                                                  manhattan_by_year$Cumulative_Shootings_By_Year,
                                                  manhattan_by_year$New_Shootings_By_Year)


# brooklyn
brooklyn_by_year <- brooklyn %>%
  mutate(Occur_Year = floor_date(Occur_Month, "year"))

brooklyn_by_year <- brooklyn_by_year %>%
  group_by(Occur_Year, BORO) %>%
  top_n(1, Cumulative_Shootings) %>%
  select(c(Occur_Year, BORO, Cumulative_Shootings)) %>%
  ungroup()

brooklyn_by_year <- brooklyn_by_year %>%
  mutate(Cumulative_Shootings_By_Year = Cumulative_Shootings)

brooklyn_by_year <- brooklyn_by_year %>%
  mutate(New_Shootings_By_Year = Cumulative_Shootings_By_Year - lag(Cumulative_Shootings_By_Year))
brooklyn_by_year$New_Shootings_By_Year <- ifelse(is.na(brooklyn_by_year$New_Shootings_By_Year), 
                                                 brooklyn_by_year$Cumulative_Shootings_By_Year,
                                                 brooklyn_by_year$New_Shootings_By_Year)

# bronx
bronx_by_year <- bronx %>%
  mutate(Occur_Year = floor_date(Occur_Month, "year"))

bronx_by_year <- bronx_by_year %>%
  group_by(Occur_Year, BORO) %>%
  top_n(1, Cumulative_Shootings) %>%
  select(c(Occur_Year, BORO, Cumulative_Shootings)) %>%
  ungroup()

bronx_by_year <- bronx_by_year %>%
  mutate(Cumulative_Shootings_By_Year = Cumulative_Shootings)

bronx_by_year <- bronx_by_year %>%
  mutate(New_Shootings_By_Year = Cumulative_Shootings_By_Year - lag(Cumulative_Shootings_By_Year))
bronx_by_year$New_Shootings_By_Year <- ifelse(is.na(bronx_by_year$New_Shootings_By_Year), 
                                              bronx_by_year$Cumulative_Shootings_By_Year,
                                              bronx_by_year$New_Shootings_By_Year)

# staten island
staten_island_by_year <- staten_island %>%
  mutate(Occur_Year = floor_date(Occur_Month, "year"))

staten_island_by_year <- staten_island_by_year %>%
  group_by(Occur_Year, BORO) %>%
  top_n(1, Cumulative_Shootings) %>%
  select(c(Occur_Year, BORO, Cumulative_Shootings)) %>%
  ungroup()

staten_island_by_year <- staten_island_by_year %>%
  mutate(Cumulative_Shootings_By_Year = Cumulative_Shootings)

staten_island_by_year <- staten_island_by_year %>%
  mutate(New_Shootings_By_Year = Cumulative_Shootings_By_Year - lag(Cumulative_Shootings_By_Year))
staten_island_by_year$New_Shootings_By_Year <- ifelse(is.na(staten_island_by_year$New_Shootings_By_Year), 
                                                      staten_island_by_year$Cumulative_Shootings_By_Year,
                                                      staten_island_by_year$New_Shootings_By_Year)

```

Here, I grouped the data by year instead of month. I also created a new column
for new shootings by year.

```{r visualize-third-analysis}

queens_by_year %>% ggplot(aes(x = Occur_Year, y = New_Shootings_By_Year)) +
  geom_line(aes(color="Queens")) +
  geom_point(aes(color="Queens")) +
  
  geom_line(data=manhattan_by_year, aes(color="Manhattan")) +
  geom_point(data=manhattan_by_year, aes(color="Manhattan")) +
  
  geom_line(data=brooklyn_by_year, aes(color="Brooklyn")) +
  geom_point(data=brooklyn_by_year, aes(color="Brooklyn")) +
  
  geom_line(data=bronx_by_year, aes(color="Bronx")) +
  geom_point(data=bronx_by_year, aes(color="Bronx")) +
  
  geom_line(data=staten_island_by_year, aes(color="Staten Island")) +
  geom_point(data=staten_island_by_year, aes(color="Staten Island")) +
  
  theme(legend.position="bottom") +
  labs(title="New York City Yearly Shootings by Borough",
       y="New Yearly Shootings",
       x="Year")

```

The graphs here are easier on the eyes. We still see that Brooklyn leads the 
pack.

```{r create-yearly-model-queens}

# queens yearly model
mod_queens_yearly <- lm(New_Shootings_By_Year ~ Occur_Year, data=queens_by_year)
summary(mod_queens_yearly)

queens_yearly_w_pred <- queens_by_year %>% mutate(pred = predict(mod_queens_yearly))

queens_yearly_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Year, y = New_Shootings_By_Year), color = "blue") + 
  geom_point(aes(x = Occur_Year, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Queens New Yearly Shootings: Actual vs Model",
       y="New Yearly Shootings",
       x="Year")

```

```{r create-yearly-model-manhattan}

# manhattan yearly model
mod_manhattan_yearly <- lm(New_Shootings_By_Year ~ Occur_Year, data=manhattan_by_year)
summary(mod_manhattan_yearly)

manhattan_yearly_w_pred <- manhattan_by_year %>% mutate(pred = predict(mod_manhattan_yearly))

manhattan_yearly_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Year, y = New_Shootings_By_Year), color = "blue") + 
  geom_point(aes(x = Occur_Year, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Manhattan New Yearly Shootings: Actual vs Model",
       y="New Yearly Shootings",
       x="Year")

```

```{r create-yearly-model-brooklyn}

# brooklyn yearly model
mod_brooklyn_yearly <- lm(New_Shootings_By_Year ~ Occur_Year, data=brooklyn_by_year)
summary(mod_brooklyn_yearly)

brooklyn_yearly_w_pred <- brooklyn_by_year %>% mutate(pred = predict(mod_brooklyn_yearly))

brooklyn_yearly_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Year, y = New_Shootings_By_Year), color = "blue") + 
  geom_point(aes(x = Occur_Year, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Brooklyn New Yearly Shootings: Actual vs Model",
       y="New Yearly Shootings",
       x="Year")

```

```{r create-yearly-model-bronx}

# bronx yearly model
mod_bronx_yearly <- lm(New_Shootings_By_Year ~ Occur_Year, data=bronx_by_year)
summary(mod_bronx_yearly)

bronx_yearly_w_pred <- bronx_by_year %>% mutate(pred = predict(mod_bronx_yearly))

bronx_yearly_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Year, y = New_Shootings_By_Year), color = "blue") + 
  geom_point(aes(x = Occur_Year, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Bronx New Yearly Shootings: Actual vs Model",
       y="New Yearly Shootings",
       x="Year")

```

```{r create-yearly-model-staten-island}

# staten_island yearly model
mod_staten_island_yearly <- lm(New_Shootings_By_Year ~ Occur_Year, data=staten_island_by_year)
summary(mod_staten_island_yearly)

staten_island_yearly_w_pred <- staten_island_by_year %>% mutate(pred = predict(mod_staten_island_yearly))

staten_island_yearly_w_pred %>% ggplot() +
  geom_point(aes(x = Occur_Year, y = New_Shootings_By_Year), color = "blue") + 
  geom_point(aes(x = Occur_Year, y = pred), color = "red") + 
  theme(legend.position="bottom",
       axis.text.x = element_text(angle=90)) +
  labs(title="Staten Island New Yearly Shootings: Actual vs Model",
       y="New Yearly Shootings",
       x="Year")

```

Linear models can now better fit the yearly data. This leads me to believe that
there are some signals that vary by month rather than by year.

# Conclusion

It appears that there is a general downward trend in the number of shootings in
New York City. However, there is a sudden increase in the number of shootings in
2020.

It appears that Brooklyn is the borough with the highest number of shootings.
Bronx comes in second. Queens has come in third, but its record is much closer
to that of Manhattan, which came in fourth. Staten Island has come in last for
the number of shootings, which makes it sound like a good place to live in.

I have heard that New York City had a dark side through scandals from the news.
I might have heard of some hearsay, like through rap songs, that Queens and 
Bronx might not be the safest places. I tried to give things a fair shake by 
trying to visualize how things truly were like, according to the data. I was
surprised that Brooklyn took the top spot instead, although Bronx took second.

I have also heard that crime could be disproportionately distributed along 
racial lines. One way I thought about addressing this is to focus on the number
of shooting incidents according to borough, ignoring the racial information
about the perpetrators and victims alike.

```{r show-session-info}
sessionInfo()
```

